---
output: 
  github_document:
    toc: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# gammit

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/m-clark/gammit.svg?branch=master)](https://travis-ci.org/m-clark/gammit)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/m-clark/gammit?branch=master&svg=true)](https://ci.appveyor.com/project/m-clark/gammit)
[![Codecov test coverage](https://codecov.io/gh/m-clark/gammit/branch/master/graph/badge.svg)](https://codecov.io/gh/m-clark/gammit?branch=master)
<!-- badges: end -->

## Introduction

The goal of gammit is to provide a set of functions to aid using `mgcv` (possibly solely) for mixed models.  Lately I've been using it in lieu of `lme4`, especially the `bam` function, for GLMM with millions of observations and multiple random effects.  It's turning out very useful in this sense (see [this post](https://m-clark.github.io/posts/2019-10-20-big-mixed-models/) for details), but I'd like some more/different functionality with the results.  Furthermore, `mgcv` just has some nice things going on for such models anyway, like the ability to add other smooth terms, alternative distributions for the target variable, etc., so I'm looking to make it easier for me to get some things I want when I use it. 

At present there are four functions: `extract_vc`, `extract_ranef`, `extract_fixed`, `summary_gamm`, and `predict_gamm.`

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("m-clark/gammit")
```

## Example

This example demonstrates the summary function with comparison to the corresponding `lme4` model.

```{r summary}
library(mgcv); library(lme4); library(gammit)

lmer_model = lmer(Reaction ~  Days + (Days || Subject), data=sleepstudy)

ga_model = gam(Reaction ~  Days + s(Subject, bs='re') + s(Days, Subject, bs='re'),
               data=sleepstudy,
               method = 'REML')

summary(lmer_model)


summary_gamm(ga_model)
```

Extract the variance components.

```{r extract_vc}
data.frame(VarCorr(lmer_model))


extract_vc(ga_model)

extract_vc(ga_model, tibble = FALSE)
```


Extract the random effects.

```{r extract_ranef}
ranef(lmer_model)


extract_ranef(ga_model)

extract_ranef(ga_model, tibble = FALSE)
```

Extract the fixed effects.

```{r extract_fixed}
fixef(lmer_model)


extract_fixed(ga_model)

extract_fixed(ga_model, tibble = FALSE)
```


## Prediction

There are a couple of ways to do prediction, and the main goal was to make it easy to use the `lme4` style to include random effects or not.  `mgcv` already has this functionality as well, so this is mostly cosmetic.  One benefit here is to provide standard errors for the prediction also.

```{r predict_gamm}
head(predict_gamm(ga_model))
```

Add standard errors.

```{r predict_gamm_se}
head(data.frame(predict_gamm(ga_model, se=T)))
```

```{r predict_gamm_fe}
compare = data.frame(
  gam_original = predict_gamm(ga_model)$prediction,
  gam_fe_only  = predict_gamm(ga_model, re_form = NA)$prediction,
  gam_fe_only2  = predict_gamm(ga_model, 
                               exclude =  c('s(Subject)', "s(Days,Subject)"))$prediction,
  lme4_fe_only = predict(lmer_model, re.form = NA))

head(compare)
```

Along with that, one can still use include/exclude for other smooth terms as above.  Unfortunately, some options do not yet work with `bam` objects, but this is to due to the functionality in `predict.gam` from `mgcv` and should change in the near future.


